\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-23 NAR_DRAFT.tex
more figure comments
figure fixes and remove comments
Revisions to dicussion

Many methods have been developed for calling somatic mutations from NGS data.
The earliest widely used somatic variant callers developed specifically for tumors, MuTect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity \citep{Koboldt2012,Cibulskis2013}.
Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \citep{Garrison2012},
joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone) \cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency-based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE) \citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and ensemble and deep learning methods (MutationSeq, BAYSIC, SomaticSeq, and SNooPer) \citep{Ding2012,Cantarel2014,Fang2015,Spinella2016}.
These methods vary in their complexity and specific focus.
But they all implicitly or explicitly assume that the rate of mutation is constant across the genome.

The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
The processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
For example, Figure~\ref{NAR-realsigfig}, shows mutation profiles from several tumors.
These real tumor mutation profiles show the diversity of profiles seen even within a single cancer.
%Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.

@@

\begin{figure}
  \begin{center}
  \includegraphics{figures/real_signatures_only.pdf}
  \end{center}
  \bkmcomment{tumor in C is probably normal breast plus APOBEC. Mention somewhere?}
  \caption{Tumor mutation profiles.
  In each panel, the x-axis corresponds to each of the 96 possible substitution types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
  \textbf{(B)} The observed mutation profile of a breast tumor used in our real data analysis \cite{Shi2018}.
  \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two breast tumors \cite{Alexandrov2019}.
   }
  \label{NAR-realsigfig}
 \end{figure}

@@

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which is known to strongly effect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \begin{array}{rcl}
  \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
  \end{array}
\end{equation}
The first two terms on the right-hand side can be estimated from the observed mutation profile (Figure ~\ref{NAR-realsigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

We model the observed mutation profile $\mathrm{S}$ (Figure~\ref{NAR-realsigfig}\textbf{A}-\textbf{D}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
In a tumor with many high-confidence observed mutations, $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile $S$.
But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
We thus model the distribution of $\mathrm{S}$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
\begin{equation}
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
\end{equation}
In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely in any context.

@@

\subsection{Implementation}
We have implemented the \batcave algorithm as an R package (batcaver).
The package leverages the Bioconductor packages \texttt{BSgenome}, \texttt{GenomicAlignments}, \texttt{VariantAnnotation}, and \texttt{SomaticSignatures} for fast and memory-efficient variant annotation and genomic context identification. 
Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.

\begin{figure}
  \begin{center}
  \includegraphics{figures/sim_signatures_only.pdf}
  \end{center}
  \caption{Tumor mutation profiles.
  In each panel, the x-axis corresponds to each of the 96 possible substitution types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{(B)} Equal proportions of signatures 1, 4, \& 5.
  \textbf{(C)} Equal proportions of signatures 1, 3, \& 5.
   }
  \label{NAR-simsigfig}
 \end{figure}

\subsection{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of allele frequencies.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) %\cite{XXX}.
\rngcomment{Put COSMIC website as a citation.}
Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}\textbf{A}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}\textbf{B}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}\textbf{C}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\citep{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
\begin{scriptsize}
\begin{verbatim}
java -Xmx24g -jar $MUTECT_JAR --analysis_type MuTect \
        --reference_sequence $ref_path \
        --dbsnp $db_snp \
        --enable_extended_output \
        --fraction_contamination 0.00 \
        --tumor_f_pretest 0.00 \
        --initial_tumor_lod -10.00 \
        --required_maximum_alt_allele_mapping_quality_score 1 \
        --input_file:normal $tmp_normal \
        --input_file:tumor $tmp_tumor \
        --out $out_path/$chr.txt \
        --coverage_file $out_path/$chr.cov
\end{verbatim}
\end{scriptsize}
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.

\subsection{Real data}
We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
We downloaded the normal and primary whole-genome AML tumor bam files from dbGaP accession number phs000159.v8.p4.
Griffith et al.\ generated a platinum set of variant calls for this tumor~\cite{Griffith2015}, which we used for our true positive dataset.
We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.

@@

To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100x whole genomes and 500x whole exomes for three different mutation profiles.
The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{NA-realsigfig}\textbf{A-C}).
We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \citep{Griffith2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
In both, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.

\begin{figure}
  \begin{center}
  \includegraphics{figures/kl_only.pdf}
  \end{center}
  \caption{Convergence of the site-specific prior to the data generating distribution. Plotted is the Kullback-Leibler divergence between the estimated and simulated profiles versus number of incorporated mutations, for \textbf{(A)} whole exomes and \textbf{(B)} whole genomes.
  }
\label{NAR-kl_fig}
\end{figure}

@@

  \begin{figure*}
    \begin{center}
    \includegraphics[width=\textwidth]{figures/fig_wes.pdf}
    \end{center}
    \caption{Variant-calling performance on both simulated and real data.
    \textbf{(A)} Precision-recall curves and \textbf{(B)} receiver operating characteristic curves for different mutation signatures.
    \textbf{(C)} and \textbf{(D)} Shaded regions show distributions of posterior probabilities for tree positive variants, and the smooth line shows the loess-smoothed relationship, from which the Integrated Calibration Index is calculated. For a perfectly calibrated caller, that curve would match the dashed y=x line. 
    \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial validation was performed \cite{Griffith2015,Shi2018}.}

  \label{NAR-wes_fig}
  \end{figure*}

@@

@@

\subsection{Tests using real tumor data}
We tested \batcave using two data sets for which deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
Both data sets were validated by deep targeted sequencing, yielding high quality false positive data.
However, \texttt{only} variants called by at least one variant caller were validated.
As a result, there is no validated measure of true/false negative calls, so we considered only precision-recall comparisons for these data. \rngcomment{Is  a correct statement?}\bkmcomment{Yes, definitely true.}

Griffith et al.\ sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>$360X and used targeted sequencing to validate nearly 200,000 mutations~\cite{Griffith2015}.
We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which is inline with previous estimates of AML mutation rates \cite{Griffith2015,Williams2018}.
For both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.995 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).

Shi et al.\ performed multi-region whole exome sequencing on 6 individual breast tumors to a mean target sequencing depth of $160\mathrm{X}$ and validated all variants identified by three different variant calling pipelines~\cite{Shi2018}.
%We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$. \rngcomment{Any comment on this result?}\bkmcomment{Noone, as far as I can tell has done a systematic calculation for TCGA. \cite{Williams2018} is only out in 2018. On the basis of tumor mutation burden studies breast cancer is one of the slowest mutators, but there is no way to go from TMB to mutation rate since it depends on depth/frequency. The one breast cancer studied in Williams 2018 the estimate at just over 1e-7. that is a whole genome whereas ours is an exome, and theirs is deeper. My argument would be that it is not out of line with previous estimates for breast cancer and it is what the data indicates.}
and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{F} and Table \ref{table:01}).
For the validated variants, MuTect and \batcave yield almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})

\begin{figure}
  \begin{center}
  \includegraphics[width=.35\textwidth]{figures/ppv_wes.pdf}
  \end{center}
  \caption{Model performance on 500X simulated whole exome. 
  Positive predictive values vs. Recall for three different probability cutoffs. 
  Note the restricted axes to improve readability of points}
\label{NAR-ppv_fig}
\end{figure}

@@

Deep sequencing is necessary to measure quantities such as intra-tumor heterogeneity \citep{Shi2018}, sub-clonal presence of driver and resistance alleles \cite{Griffith2015}, and the mutational landscape of cancer in general \cite{Zehir2017}.  
As Griffith et al., 2015~\citep{Griffith2015} show, whole genome sequencing at approx. 350X average depth requires combining calls from many variant calling algorithms, and then validating these calls via ultra-deep targeted sequencing. 
We believe that much of the need for multiple variant callers could be alleviated by improvements to the statistical model used to evaluate the evidence presented by the sequenced reads.

\batcave is a post-call variant evalutation tool that uses MuTect output and the tumor-specific mutation profile to sharpen the posterior probability assigned to SNVs.
We demonstrated here that \batcave improves classification accuracy in synthetic data, leading to better more accurate variant calls at any posterior probability threshold.
Importantly, \batcave is better calibrated than MuTect at relevant probability thresholds, allowing researchers and clinicians to make informed choices about the trade-off between precision and recall.
We are unable to study calibration in real data, because complete false positive/true negative data for real data is not available. 
Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.

@@

The \batcave algorithm focuses on the tri-nucleotide context and substitution type, which has been shown to have a large effect on site-specific mutation rates \cite{Martincorena2015,Hollstein2017}.
There are, however, many other types of \textit{genomic context} which can effect local mutation rates including replication timing, expression level, and chromosome organization \cite{Buisson2019,Schuster-Bockler2012,Pleasance2010}.
\bkmcomment{what to say here? Are we acknowledging a weakness, or just observing there are other types?}
These could be very useful in computing a local or regional mutation rate, as opposed to the genome-wide but tumor-specific mutation rate used by \batcave.

These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
At the same time, there is no way to compute true and false negative fractions for these data sets.
As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.

>>>>

Many methods have been developed for calling somatic mutations from NGS data.
The earliest widely used somatic variant callers developed specifically for tumors, MuTect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity \citep{Koboldt2012,Cibulskis2013}.
Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \citep{Garrison2012},
joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone) \cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency-based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE) \citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and ensemble and deep learning methods (MutationSeq, SomaticSeq, SNooPer, and BAYSIC) \rngcomment{Citations?}.
These methods vary in their complexity and specific focus.
But they all implicitly or explicitly assume that the rate of mutation is constant across the genome.

The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
The processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
For example, Figure~\ref{NAR-sigfig}, shows mutation profiles from several tumors.
These real tumor mutation profiles show the diversity of profiles seen even within a single cancer.
%Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.

@@

\begin{figure}
  \begin{center}
  \includegraphics{figures/signatures_only.pdf}
  \end{center}
  \bkmcomment{tumor in F is probably normal breast plus APOBEC. Mention somewhere?}
  \caption{Tumor mutation profiles.
  In each panel, the x-axis corresponds to each of the 96 possible substitution types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
  \textbf{(B)} The observed mutation profile of a breast tumor used in our real data analysis \cite{Shi2018}.
  \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two breast tumors \cite{Alexandrov2019}.
  \textbf{()} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{()} Equal proportions of signatures 1, 4, \& 5.
  \textbf{()} Equal proportions of signatures 1, 3, \& 5.
  \rngcomment{Move simulated profiles to own figure.}
   }
  \label{NAR-sigfig}
 \end{figure}

@@

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which is known to strongly effect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \begin{array}{rcl}
  \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
  \end{array}
\end{equation}
The first two terms on the right-hand side can be estimated from the observed mutation profile (Figure ~\ref{NAR-sigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

We model the observed mutation profile $\mathrm{S}$ (Figure~\ref{NAR-sigfig}\textbf{A}-\textbf{E}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
In a tumor with many high-confidence observed mutations, $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile $S$.
But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
We thus model the distribution of $\mathrm{S}$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
\begin{equation}
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
\end{equation}
In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely in any context.

@@

\subsection{Implementation}
\rngcomment{Add section about code implementation. What libraries are using, etc.?}

\subsection{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of allele frequencies.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{XXX}.
\rngcomment{Put COSMIC website as a citation.}
Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
The ``Intermediate'' profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers.
The ``Diffuse'' profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair. \rngcomment{Citations for these signature combinations.}
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\citep{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
\begin{scriptsize}
\begin{verbatim}
java -Xmx24g -jar $MUTECT_JAR --analysis_type MuTect \
        --reference_sequence $ref_path \
        --dbsnp $db_snp \
        --enable_extended_output \
        --fraction_contamination 0.00 \
        --tumor_f_pretest 0.00 \
        --initial_tumor_lod -10.00 \
        --required_maximum_alt_allele_mapping_quality_score 1 \
        --input_file:normal $tmp_normal \
        --input_file:tumor $tmp_tumor \
        --out $out_path/$chr.txt \
        --coverage_file $out_path/$chr.cov
\end{verbatim}
\end{scriptsize}
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.

\subsection{Real data}
We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
We downloaded the normal and primary whole-genome AML tumor bam files from dbGaP accession number phs000159.v8.p4.
Griffith et al.\ generated a platinum set of variant calls for this tumor~\cite{Griffith2015}, which we used for our true positive dataset.
We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.
\rngcomment{Did MuTect fail to call anything we know is there?}

@@

To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100x whole genomes and 500x whole exomes for three different mutation profiles.
The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{XXX}\textbf{A-C}).
We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \cite{Griffiths2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
In both, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.

\begin{figure}
  \begin{center}
  \includegraphics{figures/kl_only.pdf}
  \end{center}
  \caption{Convergence of the site-specific prior to the data generating distribution. Plotted is the Kullback-Leibler divergence between the estimated and simulated profiles versus number of incorporated mutations, for \textbf{(A)} whole exomes and \textbf{(B)} whole genomes.
  \rngcomment{Zoom panel B into relevant region of plot ($\sim$1000 mutations). What is notation (target || prior)?}}
\label{NAR-kl_fig}
\end{figure}

@@

  \begin{figure*}
    \begin{center}
    \includegraphics{figures/fig_wes.pdf}
    \end{center}
    \caption{Variant-calling performance on both simulated and real data.
    \textbf{(A)} Precision-recall curves and \textbf{(B)} receiver operating characteristic curves for different mutation signatures.
    \rngcomment{Make labels for \batcave and MuTect in A much bigger. Add similar labels to B.}
    \rngcomment{Move "Synthetic 500x whole exomes" to above A\&C. Add similar label for B\&D}
    \textbf{(C)} and \textbf{(D)} Shaded regions show distributions of posterior probabilities for tree positive variants, and the smooth line shows the loess-smoothed relationship, from which the Integrated Calibration Index is calculated. For a perfectly calibrated caller, that curve would match the dashed y=x line. 
    \rngcomment{Remove dark edges of distributions. Make diagonal line dashed.}
    \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial validation was performed
    \rngcomment{Label plots with what type of cancer they are, plus depth.}.}
  \label{NAR-wes_fig}
  \end{figure*}

@@

\begin{figure}
  \begin{center}
  \includegraphics{figures/ppv_wes.pdf}
  \end{center}
  \caption{Calibration of posterior probability. 
  For realistic thresholds, posterior probabilities are more accurately calibrated for \batcave than MuTect, no matter how concentration the mutation profile is.}
\label{NAR-ppv_fig}
\end{figure}

@@

\subsection{Tests using real tumor data}
We tested \batcave using two data sets for which deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
We considered only precision-recall comparisons for these data, because the validation procedure yields true positives but not comprehensive true negatives. \rngcomment{Is  a correct statement?}

Griffith et al.\ sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>$360X and used targeted sequencing to validate nearly 200,000 mutations~\cite{Griffith2015}.
We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which would be considered normal in AML. \rngcomment{Really need a citation for this!}
For both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.995 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).

Shi et al.\ performed multi-region whole exome sequencing on 6 individual breast tumors to a mean target sequencing depth of $160\mathrm{X}$ and validated all variants identified by three different variant calling pipelines~\cite{Shi2018}.
%We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$. \rngcomment{Any comment on this result?}
and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{F} and Table \ref{table:01}).
For the validated variants, MuTect and \batcave yield almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})

@@

\rngcomment{I worry that this bit about very high depth sequencing is going to be problematic in review, because it's not a problem we're solving. We're helping, but not making huge changes.}
%Deep sequencing is necessary to measure intra-tumor heterogeneity \citep{Shi2018}, identify subclonal driver and resistance alleles \cite{Griffith2015}, and probe the mutational landscape of cancer in general \cite{Zehir2017}.  
%Current approaches to very high depth next-generation sequencing \cite{Griffith2015}, however, require combining calls from multiple variant calling algorithms and then validating these calls via ultra-deep sequencing.
%We believe that much of the need for multiple variant callers could be alleviated by improvements to the statistical model used to evaluate the evidence presented by the sequenced reads.

\batcave is an algorithm that leverages the biology of individual tumor mutation profiles to improve identification of low allelic frequency somatic variants.
Our current implementation is built on MuTect, one of the most widely used somatic variant callers.
\batcave improves on the classification accuracy of MuTect in synthetic data (Fig.~\ref{NAR-wes_fig}A-D and Table~\ref{table:01}) across the entire range of recall and specificity.
Moreover, \batcave is better calibrated than MuTect at relevant posterior probability thresholds (\ref{NAR-ppv_fig}), allowing researchers and clinicians to make informed choices about the trade-off between precision and recall.
For real data, testing on validated calls shows that \batcave does not degrade performance for variants that are relatively easy to identify (Fig.~\ref{NAR-wes_fig}E\&F).
The \batcave algorithm can thus be included in a wide variety of sequencing pipelines.

The \batcave algorithm incorporates genomic context into the probabilistic model for variant calling.
Our current implementation focuses on trinucleotide context, which is known to have a large effect on local mutation rates \cite{Martincorena2015,Hollstein2017}.
There are, however, many other aspects of genomic context that can affect local mutation rates, including replication timing, expression level, and chromatin organization \cite{Buisson2019,Schuster-Bockler2012,Pleasance2010}. 
\rngcomment{Move individual citations to the factors they support.}
These additional forms of context could also be incorporated into the \batcave mutation prior.
In fact, they may be simpler to incorporate than trinucleotide context, because their effect on the prior may vary less among tumors. \rngcomment{Can we back this assertion up?} 
In the long run, we believe that incorporating more tumor biology into variant calling models will often lead to improved performance.

\rngcomment{Got here...}

@@

These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
At the same time, there is no way to compute true and false negative fractions for these data sets.
As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
%We are unable to study calibration in real data, because complete false positive/true negative data for real data is not available. 
%Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.
